---
title: "ALS_Dream_Challenge: Data_Preprocessing"
author: "Benedict Guzman"
date: "November 26, 2018"
output: html_document

---

<br>

---

```{r setdefault, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r include=FALSE, results='hide'}
library(tidyverse)
library(RCurl)
library(ggplot2)
library(ggfortify)
library(dplyr)
library(mice)
```

###**I. Introduction**

Amyotrophic lateral sclerosis (ALS) or also known as Lou Gehrig's Disease is a fatal neurodegnerative disease that kills the motor neurons controlling voluntary muscles. Subsequently, the progressive loss of these motor neurons lead to muscle atrophy and weakening and paralysis. On average, individuals afflicted with ALS live only 3-5 years upon symptom onset. It is very important to note, however, the prognosis and survival rate is heteregenous from individual to individual. Perhaps, two of the most well-known personalities who had ALS are Lou Gehrig (Baseball Player) and Stephen Hawking (Theoretical Physicist). Gehrig is a great example of a fast progressor as he died only after 2 years ofALS symptom onset. On the other hand of the spectrum,  Hawking battled the disease 49 years before he passed away just this year, 2018. 

Although it has been known for over 150 years, the biological processess and treatments for ALS has not been well-established. In fact, the only drug created for ALS is Riluzole. But even administration of this drug prolongs the survival of the patient by only a few months. What makes understanding of this disease more complicated is its aformentioned heterogenity in patients. 

To tackle this conundrum, in 2012, the DREAM-Phil Bowen ALS Prediction Prize4Life Challenge was established. Specifically, it invited participants to create algorithms to predict disease progression of ALS patients based on data within the three months after the start of the clinical observation. In 2015, a similar contest was created: the DREAM ALS Stratification Prize4Life Challenge. This time around, the participants were asked to stratify the patients in separate clusters basing on different demographic and medical characteristics that were collected within the 3 months of the start of the clinical observation. From there, the algorithm must only use 6 variables to predict disease progression by calculating the ALSFRS slope within the time frame of 3-12 months after the start of the clinical observation (1st Subchallenge). Another subchallenge was to again pick only 6 variables from each of the clusters and predict survival within the same time frame as the 1st Subchallenge requires. The datasets that can be used were from ALS Registries (data collected from Ireland and Italy as part of clinical visits of ALS patients) or from the Pooled Resource Open-Access ALS Clinical Trials (PRO-ACT) platform (data collected from over 10,000 ALS patients from 23 completed clincal trials).

As a group, we have decided to perform the first Subchallenge (ALS Progression), and the dataset that we will use is the PRO-ACT. The analysis will be divided into 3 files: Data Pre-Processing, Cluster and Feature Selector, and Data Training and Testing and Evaluation.

<br>

###**II. Data Pre-processing**

The PRO-ACT is composed of 12 different datasets, ranging from demographics to lab values. Our goal was to clean these and combine into them into 1 dataset. To perform this task, we used R statistical environment (v3.2.1) and its dpylr and mice library. In addition, we devised and adhered to the following methods:

1. Select data values that are within 90 days of the start of the clinical observation (since the challenge requires data that are within 90 days for prediction)

2. Remove variables that have more than 75% missingness 

3. Computed the mean for variables that have multiple values for the same patient (e.g. glucose level at different time points) 

4. Only the ALSFRS_Total (ALSRS_R and its scoring catergories have been eliminated because many patients were missing these scoring variables) and ALSFRS_Delta were used to calculate the ALSFRS_Slope. Specifically, the 2 points that were used was the baseline ALSFRS_Total score (ALSFRS_Delta=0) and the last ALSFRS_Total score before ALSFRS_Delta was greater than >90 days. 

5. Binarize the values of variables that are originally factors into zeroes and ones (e.g. Sex: Male=1, Female=0)

6. Impute missing data (NA values) using the Multivariate Imputation by Chained Equations (MICE) via cart method in the mice library

In the end, we were able to condense all of the information into a dataset containing 3140 patients and their 44 different features. 

```{r echo=TRUE, message=FALSE, warning=FALSE}
#original 10 datasets from PRO-ACT
ALSFRS_R <- read_csv("alsfrs.csv")
ALSHIS <- read_csv("AlsHistory.csv")
DEMOG <- read_csv("demographics.csv")
DEATH <- read_csv ("DeathData.csv")
FAMHIS <- read_csv("FamilyHistory.csv")
FVC <- read_csv("Fvc.csv")
LABS <- read_csv("Labs.csv")
RILUZOLE <- read_csv("Riluzole.csv")
SVC <- read_csv("Svc.csv")
TRTMENT <- read_csv("Treatment.csv")
VITALS <- read_csv("VitalSigns.csv")
```

Notice that there are only 10 datasets included here because upon inital inspection, both datasets entitled "adverse events" and "concomitant medications" are more than 75% missingnesss in their data.

<br>

***A. ALSFRS*** 

```{r eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
#remove rows that have data that were collected more than 90 days and eliminate the ALSFRS_R Score because the ALSFRS slope will be only calculated using the ALSFRS_Total. Also, remove the other variables pertaining to the calculation of the ALSFRS_R Score
ONE <- ALSFRS_R %>% replace_na(list(ALSFRS_Total=0)) %>% filter(ALSFRS_Delta<91, ALSFRS_Total!=0) %>% select(-ALSFRS_R_Total, -R_1_Dyspnea, -R_2_Orthopnea, -R_3_Respiratory_Insufficiency, -Mode_of_Administration, -ALSFRS_Responded_By) 
#remove columns with more than 75% NA's
ONE_Filtered <- ONE[, -which(colMeans(is.na(ONE)) > 0.75)]
#mutate a new column that calculates the slope of ALSFRS_Total and ALSFRS_Delta per subject id
ALSFRS_slope <- ONE %>% group_by(subject_id) %>% summarize(ALSFRS_Slope=(last(ALSFRS_Total)-first(ALSFRS_Total))/(last(ALSFRS_Delta)-first(ALSFRS_Delta))) %>% select(ALSFRS_Slope)
#calculate the mean foreach of the different variables for each patient
ALSFRS_mean <- ONE %>% group_by(subject_id) %>% summarise_each(funs(mean(., na.rm = TRUE)), Q1_Speech:ALSFRS_Total)
#combine the calculated slope and the calculated means into one data frame. 
ONE_COMBINED<- cbind(ALSFRS_mean, ALSFRS_slope)
#remove subject_ids with only one ALSFRS_total score because slope cannot be calculated. Also, replace NA values in the other variables into 0.
ONE_COMBINED <- ONE_COMBINED %>% drop_na(ALSFRS_Slope) %>% replace(is.na(.), 0)
```

<br>

***B. ALSFRS and ALS History***

```{r message=FALSE, warning=FALSE}
#mutate new columns for specific site of onset, for NA's treat as False
TWO <- ALSHIS %>% select(subject_id, Site_of_Onset, Onset_Delta, Site_of_Onset___Bulbar
, Site_of_Onset___Limb, Site_of_Onset___Limb_and_Bulbar
) %>% mutate(Bulbar_Onset=ALSHIS$Site_of_Onset=="Onset: Bulbar", Bulbar_Numerical=ALSHIS$Site_of_Onset___Bulbar==1, Limb_Numerical=ALSHIS$Site_of_Onset___Limb==1, Limb_Onset=ALSHIS$Site_of_Onset=="Onset: Limb", Limb_Bulbar_Onset=ALSHIS$Site_of_Onset=="Onset: Limb and Bulbar", Limb_Bulb_Numerical=ALSHIS$Site_of_Onset___Limb_and_Bulbar==1, Spine_Onset=ALSHIS$Site_of_Onset=="Onset: Spine", Other_Onset=ALSHIS$Site_of_Onset=="Onset: Other") %>% replace_na(list(Bulbar_Onset = FALSE, Limb_Bulbar_Onset = FALSE, Limb_Onset = FALSE, Spine_Onset = FALSE, Other_Onset = FALSE, Bulbar_Numerical = FALSE, Limb_Numerical = FALSE, Limb_Bulb_Numerical =FALSE)) %>% select (-Site_of_Onset, -Site_of_Onset___Bulbar,-Site_of_Onset___Limb, -Site_of_Onset___Limb_and_Bulbar) 

#convert true or false into 0 and 1, 1 being true and 0 as being false
cols <- sapply(TWO, is.logical)
TWO[,cols] <- lapply(TWO[,cols], as.numeric)
TWO_FILTERED <- TWO %>% group_by(subject_id) %>%  summarise_each(funs(mean(., na.rm = TRUE)), Onset_Delta:Other_Onset)
TWO_BINARY <- TWO_FILTERED %>% select(-subject_id, -Onset_Delta) %>% apply(2, function(x) ifelse(x > 0, 1, x))
TWO_names <- TWO_FILTERED %>% select(subject_id, Onset_Delta)
TWO_FILTERED <- cbind(TWO_names, TWO_BINARY)
#use left join to combine the als history data with the ALSFRS score. Note left_join was used because patients with ALSFRS score will be considered here
TWO_COMBINED <- ONE_COMBINED %>% left_join(TWO_FILTERED, by="subject_id") 

#assign the onset site as numerical variables
TWO_COMBINED <- TWO_COMBINED %>% mutate(Bulbar_Onset_=rowSums(TWO_COMBINED[,17:18]),Limb_Onset_=rowSums(TWO_COMBINED[,19:20]), Limb_Bulbar_=rowSums(TWO_COMBINED[,21:22]), check=rowSums(TWO_COMBINED[,17:24]))
TWO_COMBINED <- TWO_COMBINED %>% mutate(Limb_Bulbar_Check=TWO_COMBINED$check==2)
TWO_COMBINED$Limb_Bulbar_Check <- as.integer(as.logical(TWO_COMBINED$Limb_Bulbar_Check))
TWO_COMBINED <- TWO_COMBINED %>% select(-check)
TWO_COMBINED  <- TWO_COMBINED %>% mutate(Limb_Bulbar_Onset_=rowSums(TWO_COMBINED[,27:28]), Site_of_Onset=0)
TWO_COMBINED$Bulbar_Onset_[TWO_COMBINED$Limb_Bulbar_Check == 1] <- 0
TWO_COMBINED$Limb_Onset_[TWO_COMBINED$Limb_Bulbar_Check == 1] <- 0
TWO_COMBINED <- TWO_COMBINED %>% select(-Bulbar_Onset, -Limb_Bulbar_Onset, -Limb_Onset, -Bulbar_Numerical, -Limb_Bulb_Numerical, -Limb_Numerical, -Limb_Bulbar_, -Limb_Bulbar_Check) %>% filter(subject_id!="657543")
#Spine Onset=4
#Other Onset=5
#Bulbar Onset=1
#Limb Onset=2
#Limb Bulbar Onset =3
TWO_COMBINED$Site_of_Onset[TWO_COMBINED$Spine_Onset == 1] <- 4
TWO_COMBINED$Site_of_Onset[TWO_COMBINED$Other_Onset == 1] <- 5
TWO_COMBINED$Site_of_Onset[TWO_COMBINED$Bulbar_Onset_ == 1] <- 1
TWO_COMBINED$Site_of_Onset[TWO_COMBINED$Limb_Onset_ == 1] <- 2
TWO_COMBINED$Site_of_Onset[TWO_COMBINED$Limb_Bulbar_Onset_== 1] <- 3
TWO_COMBINED <-  TWO_COMBINED %>% select(-Spine_Onset,	-Other_Onset,	-Bulbar_Onset_,	-Limb_Onset_,	-Limb_Bulbar_Onset_)
TWO_COMBINED$Site_of_Onset <- as.numeric(TWO_COMBINED$Site_of_Onset)
```

<br>

***C.ALSFRS, ALS History, and Riluzole use***

```{r}
#combine ALSFRS, ALS History, and Riluzole
THREE_COMBINED <- TWO_COMBINED %>% left_join(RILUZOLE, by = "subject_id") %>% select(-Riluzole_use_Delta) %>% replace_na(list(Subject_used_Riluzole="No"))

#Convert Riluzole use as 1 and Not using Riluzole as 0. 
THREE_COMBINED$Subject_used_Riluzole <- factor(THREE_COMBINED$Subject_used_Riluzole, levels = c("Yes", "No"), c(labels = "1", "0"))
```

<br>

***D.ALSFRS, ALS History, Riluzole use, and Death days***

```{r}
#Since the goal of this analysis is to predict disease  progression, all of the subjects alive within the 90 days of the start of the ALS trial will be retained. Patients who died before that will be removed from the analyisis
FOUR <-DEATH %>% mutate(Died= DEATH$Death_Days<91 ) %>% select(subject_id, Died)

#replace NA values as false and combine with the dataset created from the chunk above
FOUR_COMBINED <- THREE_COMBINED %>% left_join(FOUR, by = "subject_id") %>% replace_na(list(Died = FALSE)) %>% filter(Died==FALSE) %>% select(-Died)
```

<br>

***E.ALSFRS, ALS History, Riluzole use, Death days, and Treatment***

```{r}
#combine ALSFRS, ALS History, Riluzole use, Death days, and Treatment. NA values would be considered as no treatment provided, and would be labeled as "None". 
FIVE_COMBINED <- FOUR_COMBINED %>% left_join(TRTMENT, by = "subject_id") %>% select(-Treatment_Group_Delta) %>% replace_na(list(Study_Arm = "None"))

#The 3 different groups are converted into factors and labeled as numerical values 2: Placebo, 1: Active, and 0: None
FIVE_COMBINED$Study_Arm <- factor(FIVE_COMBINED$Study_Arm, levels=c("Placebo", "Active", "None"), labels = c("2", "1","0"))
```

<br>

***F. ALSFRS, ALS History, Riluzole use, Death days, Treatment, and Demographics***

```{r}
#remove date of birth and demographics delta variables
SIX <-  DEMOG %>% select (-Demographics_Delta, -Date_of_Birth) 

#remove columns that have 75% missingness. In this, only age, caucasian race, and gender are left. 
SIX_FILTERED <- SIX[, -which(colMeans(is.na(SIX)) > 0.75)]

#In this case, most of the subjects in the dataset are caucsian, and therefore, the race will be binary: caucasian or not caucasian. It will be labeled as 1 for Caucasians and 0 for Non-Caucasians. Note that non-caucasian individuals could be of any race (hispanic, asian, etc.)
SIX_FILTERED_ <- SIX_FILTERED %>% replace_na(list(Race_Caucasian = FALSE))

#convert gender into numerical datatype: Male would be assigned as 1 and Female is assigned as 0. 
SIX_FILTERED_$Sex <- factor(SIX_FILTERED_$Sex, levels=c("Male", "Female"), labels = c("1", "0"))

#Combine with dataset from previous chunk
SIX_COMBINED <- FIVE_COMBINED %>% left_join(SIX_FILTERED_, by = "subject_id")
```

<br>

***G. ALSFRS, ALS History, Riluzole use, Death days, Treatment, Demographics, and SVC***

```{r message=FALSE, warning=FALSE}
#retain values that were collected within 90 days
SEVEN <- SVC %>% filter(Slow_vital_Capacity_Delta<91)

#remove columns/variables with 75% missingness
SEVEN <-  SEVEN[, -which(colMeans(is.na(SEVEN)) > 0.75)]

#since there are multiple values for the same subject in the variables, the mean is calculated.
SEVEN_FILTERED<- SEVEN %>% group_by(subject_id) %>%  summarise_each(funs(mean(., na.rm = TRUE)), Subject_Liters_Trial_1:Slow_vital_Capacity_Delta)

#combine with dataset from previous chunk
SEVEN_COMBINED <- SIX_COMBINED %>% left_join(SEVEN_FILTERED, by = "subject_id")%>% select(-Subject_Liters_Trial_1, -pct_of_Normal_Trial_1, -Slow_vital_Capacity_Delta)
```

<br>

***H. ALSFRS, ALS History, Riluzole use, Death days, Treatment, Demographics, SVC, and FVC***

```{r message=FALSE, warning=FALSE}
#retain values that were collected within 90 days
EIGHT <- FVC %>% filter(Forced_Vital_Capacity_Delta<91) %>% select(-Forced_Vital_Capacity_Units)

#remove columns/variables with 75% missingness
EIGHT <-  EIGHT[, -which(colMeans(is.na(EIGHT)) > 0.75)]

#since there are multiple values for the same subject in the variables, the mean is calculated.
EIGHT_FILTERED<- EIGHT %>% group_by(subject_id) %>%  summarise_each(funs(mean(., na.rm = TRUE)), Subject_Liters_Trial_1:Subject_Normal)

#combine with dataset from previous chunk
EIGHT_COMBINED <- SEVEN_COMBINED %>% left_join(EIGHT_FILTERED, by = "subject_id")
```

<br>

***H. ALSFRS, ALS History, Riluzole use, Death days, Treatment, Demographics, SVC, FVC, and Family History***

```{r message=FALSE, warning=FALSE}
#remove Family_History_Delta variable as all values are zero
NINE <- FAMHIS %>% select(-Family_History_Delta)

#remove columns/variables with 75% missingness
NINE <-  NINE[, -which(colMeans(is.na(NINE)) > 0.75)]

#create a new column with boolean values if the patient has a relative who had/has any type of neurological disease including ALS
NINE_FILTERED <- NINE %>% mutate(Fam_Hist_Neuro_Dis=is.na(NINE$Neurological_Disease)) %>% select(-Neurological_Disease)

#Note that 1 means that the patient has/had no relative witha neurological disease (including ALS) and 0 means that the patient has/had a relative with a neurological disease
cols <- sapply(NINE_FILTERED, is.logical)
NINE_FILTERED[,cols] <- lapply(NINE_FILTERED[,cols], as.numeric)
NINE_COMBINED <- EIGHT_COMBINED %>% left_join(NINE_FILTERED, by = "subject_id") %>% replace_na(list(Fam_Hist_Neuro_Dis = 1))
```


<br>

***I. ALSFRS, ALS History, Riluzole use, Death days, Treatment, Demographics, SVC, FVC, Family History, and Vitals***

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
#filter data that were collected up to 90 days only
TEN <- VITALS %>% filter(Vital_Signs_Delta<91)

#remove variables that have more than 75% missingness and eliminate the columns with the unit labels
#blood_pressure units is mmHg
#height units is inches
#weight units is kg
#pulse units is bpm
#temperature is Celcius
#respiratory rate is resp/min
TEN <-  TEN[, -which(colMeans(is.na(TEN)) > 0.75)] %>% select(-Blood_Pressure_Diastolic_Units, -Blood_Pressure_Systolic_Units, -Pulse_Units, -Respiratory_Rate_Units, -Temperature_Units, -Weight_Units)

#Take the mean for each of the different vital signs for each patient
TEN_FILTERED<- TEN %>% group_by(subject_id) %>%  summarise_each(funs(mean(., na.rm = TRUE)), Blood_Pressure_Diastolic:Vital_Signs_Delta)

#combine with dataset from previous chunk
TEN_COMBINED <- NINE_COMBINED %>% left_join(TEN_FILTERED, by = "subject_id")
```

<br>

***J. ALSFRS, ALS History, Riluzole use, Death days, Treatment, Demographics, SVC, FVC, Family History, Vitals, and Laboratory Tests***

```{r}
#filter data that were collected up to 90 days only and select lab values with more than 75% of the patient population having tested for it
ELEVEN <- LABS %>% filter(Laboratory_Delta<91)%>% group_by(Test_Name) %>% tally() %>% filter(n>22000)

#there are only 9 lab tests to be retained
ELEVEN
```
```{r message=FALSE, warning=FALSE}
#create columns for each of the 9 tests determined above. For NA values, replace them with zeroes.
ELEVEN_FILTERED <- LABS %>% filter(Test_Name %in% c("ALT(SGPT)","AST(SGOT)", "Creatinine", "Glucose", "Hematocrit", "Hemoglobin", "Platelets", "Red Blood Cells (RBC)", "White Blood Cell (WBC)"), Laboratory_Delta<91, Test_Result!="-") %>% mutate(row = row_number()) %>% spread(Test_Name, Test_Result) %>% group_by(subject_id) %>% replace_na(list(`ALT(SGPT)` = 0,`AST(SGOT)`= 0, Creatinine= 0, Glucose=0, Hematocrit=0, Hemoglobin=0, Platelets=0, `Red Blood Cells (RBC)`=0, `White Blood Cell (WBC)`= 0)) %>% select(-Test_Unit, -row, -Laboratory_Delta)
ELEVEN_FILTERED <- ELEVEN_FILTERED%>% filter()

#coerce each of the variables into numeric datatype
ELEVEN_FILTERED$`ALT(SGPT)`<- as.numeric(ELEVEN_FILTERED$`ALT(SGPT)`)
ELEVEN_FILTERED$`AST(SGOT)`<- as.numeric(ELEVEN_FILTERED$`AST(SGOT)`)
ELEVEN_FILTERED$Creatinine <- as.numeric(ELEVEN_FILTERED$Creatinine)
ELEVEN_FILTERED$Glucose <- as.numeric(ELEVEN_FILTERED$Glucose)
ELEVEN_FILTERED$Hematocrit <- as.numeric(ELEVEN_FILTERED$Hematocrit)
ELEVEN_FILTERED$Hemoglobin<- as.numeric(ELEVEN_FILTERED$Hemoglobin)
ELEVEN_FILTERED$Platelets <- as.numeric(ELEVEN_FILTERED$Platelets)
ELEVEN_FILTERED$`Red Blood Cells (RBC)` <- as.numeric(ELEVEN_FILTERED$`Red Blood Cells (RBC)`)
ELEVEN_FILTERED$`White Blood Cell (WBC)` <- as.numeric(ELEVEN_FILTERED$`White Blood Cell (WBC)`)

#for lab tests with multiple values for the same patient, calculate the mean
ELEVEN_FILTERED <- ELEVEN_FILTERED %>% rename(ALT= `ALT(SGPT)`, AST =`AST(SGOT)`, RBC= `Red Blood Cells (RBC)`, WBC=`White Blood Cell (WBC)`) %>% summarise_each(funs(mean(.[.!=0])),ALT:WBC)

#combine with the dataset created in the previous chunk
ELEVEN_COMBINED <- TEN_COMBINED %>% left_join(ELEVEN_FILTERED, by = "subject_id")
```

<br>

***K. Data Imputation***

```{r}
ALS_RAW <-  ELEVEN_COMBINED[, -which(colMeans(is.na(ELEVEN_COMBINED)) > 0.25)]
#remove rows that have 75% missing data
#remove rows with zeroes in the lab values
ALS_RAW<- ALS_RAW %>% filter(AST!=0, ALT!=0, Creatinine!=0, Glucose!=0, Hematocrit!=0, Hemoglobin!=0, Hematocrit!=0,Platelets!=0, RBC!=0, WBC!=0)

#coerce the remaining variables that are not numerical datatypes
ALS_RAW$Subject_used_Riluzole  <- as.numeric(ALS_RAW$Subject_used_Riluzole)
ALS_RAW$Study_Arm <- as.numeric(ALS_RAW$Study_Arm)
ALS_RAW$Sex <- as.numeric(ALS_RAW$Sex)
ALS_RAW$Race_Caucasian <- as.numeric(ALS_RAW$Race_Caucasian)
str(ALS_RAW)
```
```{r}
#perform data imputation using MICE via cart method
tempData <- mice(ALS_RAW,m=5,maxit=10,meth='cart',seed=500)
completedData <- complete(tempData,1)
ALS_FINAL <- completedData
dim(ALS_FINAL)
write.csv(ALS_FINAL, "ALS_FINAL_Re.csv")
